---
output: github_document
---

<!-- WARNING: README.md is generated from README.Rmd; edit that
     instead, then use rmarkdown::render("README.Rmd") to regenerate -->

```{r, echo=FALSE}
knitr::opts_chunk$set(
  fig.path = "man/figures/README-"
)
```

# the regressinator

The regressinator is a pedagogical tool for conducting simulations of regression
analyses and diagnostics. It can:

* Simulate populations with predictor variables from arbitrary distributions
* Simulate response variables that are function of the predictor variables plus
  error, or are drawn from a distribution related to the predictors
* Given a model fit to simulated data, generate new simulated data based on the
  model fit
* Facilitate lineup plots comparing diagnostics on the fitted model to
  diagnostics where all model assumptions are met.

Or, in other words: We can specify a population with a known relationship
between predictors in response, simulate data from that population, fit a model
to that data, and simulate data from the model. If the fitted model is
misspecified, we can simulate data from the model (where it is *correctly*
specified), and make a lineup to compare the fitted model's diagnostics to null
diagnostics.

Here's a simple regression example:

```{r example-regression-lineup, fig.width=7, fig.height=6}
library(regressinator)
library(ggplot2)

# true relationship is nonlinear
pop <- population(
  x = predictor("runif", min = -5, max = 5),
  y = response(10 + 0.7 * x**2, family = gaussian(), error_scale = 2)
)

nonlin_sample <- pop |>
  sample_x(n = 20) |>
  sample_y()

fit <- lm(y ~ x, data = nonlin_sample)

diagnose_model(fit) |>
  ggplot(aes(x = x, y = .resid)) +
  geom_point() +
  facet_wrap(~ .sample) +
  labs(x = "x", y = "Residual")
```

Each of the 20 residual plots represents a simple regression fit. One of the
regressions was fit to the simulated data with a nonlinear relationship; the
others were fit to data simulated *from the linear model*, using the same model.
(That is, they were simulated from `fit`, and then the same model was refit to
them.) So one of these residual plots shows the residuals of a linear model fit
to nonlinear data, and the others show the residuals of a linear model fit to
*linear* data.

The premise of the regressinator is that this kind of simulation can be a
valuable teaching tool when students are learning to interpret regression
diagnostics and determine how different kinds of misspecification might affect
model fits.

Check the Get Started guide for more detail.

## Inspirations

Visual inference -- hiding a plot of real data among "null plots" -- has been
around for quite a while, and I certainly did not invent this idea myself. For
example:

* Buja, A., Cook, D., Hofmann, H., Lawrence, M., Lee, E. K., Swayne, D. F., &
  Wickham, H. (2009). [Statistical inference for exploratory data analysis and
  model diagnostics](https://doi.org/10.1098/rsta.2009.0120). *Philosophical
  Transactions of the Royal Society A*, 367(1906), 4361–4383.

* Wickham, H., Cook, D., Hofmann, H., & Buja, A. (2010). [Graphical inference
  for infovis](https://doi.org/10.1109/TVCG.2010.161). *IEEE Transactions on
  Visualization and Computer Graphics*, 16(6), 973–979.

The idea of using visual inference to teach regression diagnostics is also not
new:

* Loy, A. (2021). [Bringing Visual Inference to the
  Classroom](https://doi.org/10.1080/26939169.2021.1920866). *Journal of
  Statistics and Data Science Education*, 29(2), 171-182.

The regressinator simply adds an easy-to-use framework to allow all kinds of
teaching activities to be constructed quickly. Instructors can design example to
display in lecture, or students with R experience can run interactive examples
and explore different situations. Ideally, if a student asks "But what if *[some
problem with the model]* happens?", you should be able to reply with a quick
simulation.
