---
output: github_document
---

<!-- WARNING: README.md is generated from README.Rmd; edit that
     instead, then use rmarkdown::render("README.Rmd") to regenerate -->

```{r, echo=FALSE}
knitr::opts_chunk$set(
  fig.path = "man/figures/README-"
)
```

# the regressinator

The regressinator is a pedagogical tool for conducting simulations of regression
analyses and diagnostics. It can:

* Simulate populations with predictor variables from arbitrary distributions
* Simulate response variables that are function of the predictor variables plus
  error, or are drawn from a distribution related to the predictors
* Given a model, simulate from the population sampling distribution of that
  model's estimates
* Given a model fit to simulated data, generate new simulated data based on the
  model fit
* Facilitate lineup plots comparing diagnostics on the fitted model to
  diagnostics where all model assumptions are met.

Or, in other words: We can specify a population with a known relationship
between predictors in response, simulate data from that population, fit a model
to that data, and simulate data from the model. If the fitted model is
misspecified, we can simulate data from the model (where it is *correctly*
specified), and make a lineup to compare the fitted model's diagnostics to null
diagnostics. And we can obtain the population sampling distribution of the
misspecified model's estimates.

Here's a simple regression example:

```{r example-regression-lineup, fig.width=7, fig.height=6, fig.retina=2}
library(regressinator)
library(ggplot2)

# true relationship is nonlinear
pop <- population(
  x = predictor("runif", min = -5, max = 5),
  y = response(10 + 0.7 * x**2, family = gaussian(), error_scale = 2)
)

nonlin_sample <- pop |>
  sample_x(n = 20) |>
  sample_y()

fit <- lm(y ~ x, data = nonlin_sample)

model_lineup(fit) |>
  ggplot(aes(x = x, y = .resid)) +
  geom_point() +
  facet_wrap(~ .sample) +
  labs(x = "x", y = "Residual")
```

Each of the 20 residual plots represents a simple regression fit. One of the
regressions was fit to the simulated data with a nonlinear relationship; the
others were fit to data simulated *from the linear model*, using the same model.
(That is, they were simulated from `fit`, and then the same model was refit to
them.) `model_lineup()`, by default, uses `broom::augment()` to obtain the data
and residuals from each fit. So one of these residual plots shows the residuals
of a linear model fit to nonlinear data, and the others show the residuals of a
linear model fit to *linear* data.

Similarly, we can quickly obtain samples from the population sampling
distribution, to explore the behavior of this misspecified model's parameter
estimates:

```{r sampling-distribution}
fit |>
  sampling_distribution(nonlin_sample, nsim = 5)
```

Here `sampling_distribution()` defaults to using `broom::tidy()` to obtain the
coefficients and standard errors for each fit. We could use this to assess bias,
to compare the sampling distribution to standard errors and confidence
intervals, or simply to visualize variation.

The premise of the regressinator is that this kind of simulation can be a
valuable teaching tool when students are learning to interpret regression
diagnostics and determine how different kinds of misspecification might affect
model fits.

Check the Get Started guide (`vignette("regressinator")`) for more detail.

## Inspirations

Visual inference -- hiding a plot of real data among "null plots" -- has been
around for quite a while, and I certainly did not invent this idea myself. For
example:

* Buja, A., Cook, D., Hofmann, H., Lawrence, M., Lee, E. K., Swayne, D. F., &
  Wickham, H. (2009). [Statistical inference for exploratory data analysis and
  model diagnostics](https://doi.org/10.1098/rsta.2009.0120). *Philosophical
  Transactions of the Royal Society A*, 367(1906), 4361–4383.

* Wickham, H., Cook, D., Hofmann, H., & Buja, A. (2010). [Graphical inference
  for infovis](https://doi.org/10.1109/TVCG.2010.161). *IEEE Transactions on
  Visualization and Computer Graphics*, 16(6), 973–979.

* Loy, A., Follett, L., & Hofmann, H. (2016). [Variations of Q-Q Plots: The
  Power of Our Eyes!](https://doi.org/10.1080/00031305.2015.1077728) *The
  American Statistician*, 70(2), 202–214.

The idea of using visual inference to teach regression diagnostics is also not
new:

* Loy, A. (2021). [Bringing Visual Inference to the
  Classroom](https://doi.org/10.1080/26939169.2021.1920866). *Journal of
  Statistics and Data Science Education*, 29(2), 171-182.

The regressinator simply adds an easy-to-use framework to allow all kinds of
teaching activities to be constructed quickly. Instructors can design example to
display in lecture, or students with R experience can run interactive examples
and explore different situations. Ideally, if a student asks "But what if *[some
problem with the model]* happens?", you should be able to reply with a quick
simulation.

## Compared to other packages

There have been several past efforts to support pedagogical simulation and
lineup plots in R:

* [nullabor](https://cran.r-project.org/package=nullabor), which supports lineup
  plots. The regressinator uses nullabor underneath when building lineups.
* [mosaic](https://cran.r-project.org/package=mosaic), part of [Project
  MOSAIC](http://www.mosaic-web.org/), provides a simple set of functions for
  doing EDA and basic inferential statistics, including the `do()` function for
  easy simulation without loops.
* [infer](https://infer.tidymodels.org/), part of the
  [tidymodels](https://www.tidymodels.org/) framework, can conduct
  simulation-based inference for proportions, means, regression slopes, and
  other estimates commonly used in statistics courses, using only a few simple
  functions.

Unlike these packages, the regressinator provides a simple tool for specifying a
*population* and sampling from it, rather than conducting bootstrapping or
permutation on an observed dataset. This makes the regressinator suitable for,
say, exploring the properties of regression estimates and diagnostics in known
populations, but less suitable for simulation-based hypothesis testing.

Unlike infer, the regressinator does not wrap R statistical methods or provide
its own inference functions. Users must use `lm()`, `glm()`, or whatever other
methods they need for their modeling. This makes the regressinator less suitable
for introductory courses where extra complexity should be hidden away from
students, but more suitable for more advanced work: as students advance to more
complex models provided by other packages, they can use those models in the
regressinator, without any special support being required.

A useful counterpart to the regressinator might be
[rsample](https://rsample.tidymodels.org/), a general framework for methods that
resample from the observed data, such as bootstrapping. In the same way that the
regressinator supports general-purpose simulation from the population without
hard-coding specific use cases, rsample supports resampling and cross-validation
in a general way that could be used for any kind of modeling, not just models
built into the package.
