---
title: "Example linear regression diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example linear regression diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(regressinator)
library(ggplot2)
library(broom)
```

For these examples, we'll consider a simple multivariate setting where the
population relationship is nonlinear and we misspecify the model:

```{r}
nonlinear_pop <- population(
  x1 = predictor("runif", min = 1, max = 8),
  x2 = predictor("runif", min = 4, max = 12),
  y = response(0.7 + 0.8 * x1**2 + 1.2 * x2,
               family = gaussian(), error_scale = 4.0)
)

nonlinear_data <- sample_x(nonlinear_pop, n = 100) |>
  sample_y()

fit <- lm(y ~ x1 + x2, data = nonlinear_data)
```

We'll use `broom::augment()` to get diagnostic information from the fit in a
standardized data frame, making it easy to use ggplot2 to produce different
diagnostic plots.

## Residual plots

Residuals versus fitted values:

```{r}
augment(fit) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  labs(x = "Fitted value", y = "Residual")
```

To plot residuals versus the covariates, we can use `augment_longer()`, which is
like `augment()` but converts the data to long form with one row per predictor
per observation, making it easy to facet the residuals:

```{r}
augment_longer(fit) |>
  ggplot(aes(x = .predictor_value, y = .resid)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  facet_wrap(vars(.predictor_name), scales = "free_x") +
  labs(x = "Predictor", y = "Residual")
```

Using `model_lineup()`, we can compare the true residual plots to several where
the model is correctly specified. Each row gives the plots for one simulation,
and one of the fives rows (at random) is the true residual plots:

```{r, fig.height=8}
model_lineup(fit, fn = augment_longer, n = 5) |>
  ggplot(aes(x = .predictor_value, y = .resid)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(.sample), cols = vars(.predictor_name),
             scales = "free_x") +
  labs(x = "Predictor", y = "Residual")
```

## Partial residual plots

The `partial_residuals()` function fetches partial residuals in a convenient
data frame format. See its documentation for references on the use and
interpretation of partial residuals.

```{r}
partial_residuals(fit) |>
  ggplot(aes(x = predictor_value, y = partial_resid)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  facet_wrap(vars(predictor_name), scales = "free") +
  labs(x = "Predictor", y = "Partial residual")
```

We can again use `model_lineup()` to see how these compare to the partial
residuals when the model is correctly specified:

```{r, fig.height=8}
model_lineup(fit, partial_residuals, n = 5) |>
  ggplot(aes(x = predictor_value, y = partial_resid)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(.sample), cols = vars(predictor_name),
             scales = "free_x") +
  labs(x = "Predictor", y = "Partial residual")
```

## Cook's distances

The Cook's distance for an observation represents how much the model fitted
values would change if that observation were removed, scaled by the model's mean
squared error. A Cook's distance of 1 is often considered a cutoff for a highly
influential observation.

```{r}
augment(fit) |>
  ggplot(aes(x = seq_along(.cooksd), y = .cooksd)) +
  geom_col() +
  labs(x = "Row index", y = "Cook's distance")
```

Note that the Cook's distance measures changes in fitted values, *not* in $\hat
\beta$. If several predictors are collinear, a small change in an observation
may change $\hat \beta$ dramatically but not change the fitted values much.

## Residual Q-Q plots

To see the distribution of the standardized residuals, and compare it to a
normal distribution:

```{r}
augment(fit) |>
  ggplot(aes(sample = .std.resid)) +
  geom_qq() +
  geom_qq_line() +
  labs(title = "Normal Q-Q plot of standardized residuals",
       x = "Theoretical quantiles", y = "Observed quantiles")
```

It is difficult for novices to judge normality from a Q-Q plot, so a lineup can
be helpful:

```{r, fig.height=8}
model_lineup(fit) |>
  ggplot(aes(sample = .std.resid)) +
  geom_qq() +
  geom_qq_line() +
  facet_wrap(vars(.sample)) +
  labs(title = "Normal Q-Q plot of standardized residuals",
       x = "Theoretical quantiles", y = "Observed quantiles")
```

The true Q-Q plot does not stand out here, so we have little evidence for a
systematically non-normal residuals.
